---
title: "Broadband as Economic Infrastructure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Broadband as Economic Infrastructure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  eval       = FALSE,
  fig.width  = 8.888889,
  fig.height = 6.25,
  dpi        = 300
)
```

```{r libraries}
library(cori.data)
library(cori.data.qcew)
library(cori.data.fcc)
library(ruraldefinitions)
library(cori.charts)
library(dplyr)
library(ggplot2)
library(sf)
library(tigris)
library(scales)
library(stringr)

load_fonts()
update_cori_geom_defaults()
```

Broadband access is increasingly inseparable from economic participation.
Remote work, e-commerce, telemedicine, and digital services all depend on
reliable, high-speed internet — and rural areas have historically faced the
greatest gaps in coverage.

The `cori.data.fcc` package provides access to the FCC's National Broadband
Map (NBM), which tracks broadband serviceable locations at the address level.
This vignette shows how to pull county-level broadband data and connect it
to economic outcomes.

## How FCC NBM data works

The NBM measures whether locations can **receive** broadband service at a
given speed threshold — it does not measure actual adoption or usage. Key
metrics available at the county level:

| Variable | Description |
|---|---|
| `cnt_total_locations` | Total serviceable locations |
| `cnt_fiber_locations` | Locations with fiber infrastructure |
| `cnt_100_20` | Locations with 100/20 Mbps coverage |
| `cnt_25_3` | Locations with 25/3 Mbps coverage (legacy FCC threshold) |

Data is published in releases: `"D23"` (December 2023), `"J24"` (June 2024),
`"D24"` (December 2024), `"J25"` (June 2025). Use `release = "latest"` for
the most current data.

## Pulling broadband data

`get_nbm_county()` queries one county at a time. For a state or regional
analysis, loop across county FIPS codes using `lapply()`:

```{r broadband-data}
# Get all West Virginia county FIPS codes
wv_counties <- tiger_line_counties() |>
  filter(substr(GEOID, 1, 2) == "54") |>
  pull(GEOID)

# Pull broadband coverage for each county
broadband_wv <- lapply(wv_counties, function(g) {
  tryCatch(
    get_nbm_county(geoid_co = g, release = "latest"),
    error = function(e) {
      message(sprintf("Skipping %s: %s", g, conditionMessage(e)))
      NULL
    }
  )
}) |>
  bind_rows() |>
  mutate(
    pct_100_20 = cnt_100_20 / cnt_total_locations,
    pct_fiber  = cnt_fiber_locations / cnt_total_locations
  )
```

## Map: broadband coverage in West Virginia

```{r broadband-map, fig.alt = "Choropleth map of 100/20 Mbps broadband coverage across West Virginia counties"}
wv_counties_sf <- tiger_line_counties() |>
  filter(substr(GEOID, 1, 2) == "54") |>
  left_join(broadband_wv, by = c("GEOID" = "geoid"))

wv_state_sf <- tiger_line_states() |>
  filter(GEOID == "54")

fig_map <- ggplot() +
  geom_sf(
    data      = wv_state_sf,
    fill      = "#FBF8E9",
    color     = "#d0d2ce",
    linewidth = 0.6
  ) +
  geom_sf(
    data  = wv_counties_sf,
    aes(fill = pct_100_20),
    color = "#d0d2ce",
    linewidth = 0.3
  ) +
  scale_fill_cori(
    palette  = "ctg3gn",
    discrete = FALSE,
    labels   = label_percent(accuracy = 1),
    na.value = "#d0d2ce"
  ) +
  theme_cori_map() +
  labs(
    title    = "Broadband access remains uneven across West Virginia",
    subtitle = "Share of serviceable locations with 100/20 Mbps coverage, latest FCC NBM release",
    caption  = str_wrap(
      "Source: FCC National Broadband Map via cori.data.fcc. Coverage reflects serviceable locations, not adoption.",
      width = 110
    ),
    fill = "100/20 Mbps\ncoverage"
  )

fig_map
save_plot(fig_map, here::here("export/broadband_wv_map.png"))
```

## Connecting broadband to employment

Does broadband coverage correlate with stronger employment outcomes?
Joining FCC data with QCEW employment and rural classification gives
a first look:

```{r broadband-employment}
# Employment rate in WV counties
emp <- get_employment(geoids = wv_counties, years = 2023) |>
  filter(variable == "employment") |>
  select(geoid, employment = value)

pop16 <- get_population(geoids = wv_counties, years = 2023) |>
  filter(variable == "population_16plus") |>
  select(geoid, pop_16plus = value)

rural <- get_definition("cbsa", 2023) |> select(geoid, is_rural)

wv_combined <- broadband_wv |>
  left_join(emp,   by = "geoid") |>
  left_join(pop16, by = "geoid") |>
  left_join(rural, by = "geoid") |>
  mutate(emp_rate = employment / pop_16plus) |>
  filter(!is.na(emp_rate), !is.na(pct_100_20))
```

```{r broadband-scatter, fig.alt = "Scatter plot of broadband coverage vs employment rate for West Virginia counties, sized by total locations and colored by rural status"}
fig_scatter <- wv_combined |>
  ggplot(aes(
    x     = pct_100_20,
    y     = emp_rate,
    color = is_rural,
    size  = cnt_total_locations
  )) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, aes(group = 1),
              color = cori_colors["CORI Gray"]) +
  scale_color_cori(palette = "ctg2ruralnonrural") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_size_continuous(
    range  = c(1, 8),
    labels = label_comma(),
    name   = "Total locations"
  ) +
  theme_cori() +
  labs(
    title    = "Broadband coverage and employment rate, West Virginia counties",
    subtitle = "2023 data. Each dot is a county.",
    caption  = str_wrap(
      "Sources: FCC National Broadband Map; BLS QCEW; Census PEP. Rural classification: CBSA 2023.",
      width = 110
    ),
    color = NULL,
    x     = "Share of locations with 100/20 Mbps coverage",
    y     = "Covered employment rate (16+)"
  )

fig_scatter
save_plot(fig_scatter, here::here("export/broadband_employment_scatter.png"))
```

## A note on interpretation

The FCC NBM measures **availability**, not **adoption**. A location counted
as covered may not have a subscribing household. In rural areas, cost and
digital literacy barriers can mean that even where coverage exists,
actual connectivity lags behind what the map suggests. Use availability data
as a floor estimate — the actual connected share is likely lower.

For deeper analysis of broadband and economic outcomes in rural communities,
see [CORI's broadband research](https://ruralinnovation.us/our-work/research/).

## Data sources

> Federal Communications Commission. *National Broadband Map*.
> <https://broadbandmap.fcc.gov/>
>
> U.S. Bureau of Labor Statistics. *Quarterly Census of Employment and Wages*.
> <https://www.bls.gov/cew/>
>
> U.S. Census Bureau. *Population Estimates Program*.
> <https://www.census.gov/programs-surveys/popest.html>
>
> U.S. Census Bureau. *Metropolitan and Micropolitan Statistical Areas (OMB delineations)*.
> <https://www.census.gov/programs-surveys/metro-micro.html>
