---
title: "Employment Rate: A Cross-Package Story"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Employment Rate: A Cross-Package Story}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  eval       = FALSE,
  fig.width  = 8.888889,
  fig.height = 6.25,
  dpi        = 300
)
```

```{r libraries}
library(cori.data)
library(cori.data.qcew)
library(cori.data.pep)
library(ruraldefinitions)
library(cori.charts)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(tigris)
library(scales)
library(stringr)

load_fonts()
update_cori_geom_defaults()
```

Employment levels tell you how many people work somewhere. But without context
— how many people *could* work — they don't tell you much about a local labor
market's health. A county with 10,000 covered workers looks very different if
its working-age population is 20,000 versus 200,000.

This vignette builds a more meaningful measure — **covered employment as a
share of working-age population (16+)** — by combining data from two packages.

## Pulling the data

```{r data}
# Covered employment from BLS QCEW — total workers, private + government
emp <- get_employment(geography = "county", years = 2010:2023)

# Working-age population (16+) from Census PEP
pop16 <- get_population(
  geography = "county",
  years     = 2010:2023,
  variables = "population_16plus"
)

# Rural classification — CBSA-based, 2023 vintage
rural <- get_definition("cbsa", 2023) |>
  select(geoid, is_rural)
```

## Computing the rate

Both datasets are tidy, so the join is clean. We extract the relevant
variable from each, then divide.

```{r compute}
emp_wide <- emp |>
  filter(variable == "employment") |>
  select(geoid, year, employment = value)

pop_wide <- pop16 |>
  filter(variable == "population_16plus") |>
  select(geoid, year, pop_16plus = value)

emp_rate <- emp_wide |>
  inner_join(pop_wide, by = c("geoid", "year")) |>
  mutate(emp_rate = employment / pop_16plus) |>
  left_join(rural, by = "geoid") |>
  filter(!is.na(is_rural), !is.na(emp_rate))
```

## Line chart: rural vs. nonrural over time

We use an employment-weighted average so that large counties don't get
the same voice as small ones. See the
[Weighted Averages vignette](weighted-averages.html) for more on why this matters.

```{r line-chart, fig.alt = "Line chart showing employment-weighted average employment rates for rural and nonrural counties from 2010 to 2023"}
trend <- emp_rate |>
  group_by(year, is_rural) |>
  summarize(
    avg_emp_rate = weighted.mean(emp_rate, employment, na.rm = TRUE),
    .groups = "drop"
  )

title    <- "Rural counties have trailed nonrural on employment rates\nsince the Great Recession"
subtitle <- "Employment-weighted average covered employment rate, 2010\u20132023"
caption  <- str_wrap(
  "Source: BLS Quarterly Census of Employment and Wages; U.S. Census Bureau Population Estimates Program. Rural classification: CBSA 2023. Note: covered employment excludes self-employed workers.",
  width = 110
)

fig_line <- trend |>
  ggplot(aes(x = year, y = avg_emp_rate, color = is_rural)) +
  geom_line(linewidth = 1.5) +
  scale_color_cori(palette = "ctg2ruralnonrural") +
  scale_x_continuous(breaks = seq(2010, 2023, by = 2)) +
  scale_y_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  theme_cori() +
  theme(legend.position = "right", legend.direction = "vertical") +
  guides(color = guide_legend(byrow = TRUE, reverse = TRUE)) +
  labs(
    title    = title,
    subtitle = subtitle,
    caption  = caption,
    color    = NULL,
    x        = NULL,
    y        = NULL
  )

fig_line
save_plot(fig_line, here::here("export/employment_rate_line.png"))
```

## County map: where are employment rates lowest?

```{r map, fig.alt = "Choropleth map of county-level employment rates across the United States, 2023"}
counties_sf <- tiger_line_counties() |>
  shift_geometry() |>
  select(geoid = GEOID)

states_sf <- tiger_line_states() |>
  shift_geometry()

map_data <- emp_rate |>
  filter(year == 2023) |>
  select(geoid, emp_rate)

counties_map <- counties_sf |>
  left_join(map_data, by = "geoid")

fig_map <- ggplot() +
  geom_sf(
    data      = states_sf,
    fill      = "#FBF8E9",
    color     = "#d0d2ce",
    linewidth = 0.4
  ) +
  geom_sf(
    data  = counties_map,
    aes(fill = emp_rate),
    color = NA
  ) +
  scale_fill_cori(
    palette  = "ctg3gn",
    discrete = FALSE,
    labels   = label_percent(accuracy = 1),
    na.value = "#d0d2ce"
  ) +
  theme_cori_map() +
  labs(
    title    = "Employment rates vary widely across rural counties",
    subtitle = "Covered employment as share of population 16+, 2023",
    caption  = str_wrap(
      "Source: BLS QCEW; Census PEP. Rural classification: CBSA 2023.",
      width = 110
    ),
    fill = "Employment\nrate"
  )

fig_map
save_plot(fig_map, here::here("export/employment_rate_map.png"))
```

## A note on coverage

QCEW employment counts **covered** workers — those employed by businesses
that pay into unemployment insurance. Self-employed workers, independent
contractors, and some farm workers are not included. This means the
employment rate derived here is not directly comparable to the BLS official
unemployment rate, which uses a different survey methodology. It is most
useful for comparing trends and relative differences across geographies.
