---
title: "Sector Analysis: How Rural Economies Differ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sector Analysis: How Rural Economies Differ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  eval       = FALSE,
  fig.width  = 8.888889,
  fig.height = 6.25,
  dpi        = 300
)
```

```{r libraries}
library(cori.data.qcew)
library(ruraldefinitions)
library(cori.charts)
library(dplyr)
library(ggplot2)
library(ggbeeswarm)
library(scales)
library(stringr)

load_fonts()
update_cori_geom_defaults()
```

Not all employment is equal. A county where most workers are in healthcare
and local retail faces very different economic dynamics than one anchored in
manufacturing or professional services. Sector composition shapes everything
from wage levels to resilience against downturns to sensitivity to remote work.

`cori.data.qcew` offers two ways to cut employment by sector:

- **BLS supersectors** — the 11 standard industry groupings published by
  the Bureau of Labor Statistics (natural resources, construction,
  manufacturing, etc.)
- **CORI super-sectors** — a CORI-designed aggregation into 3 categories
  based on whether industries primarily face local or national/global demand

## CORI super-sectors

CORI's classification, loosely derived from Eckert (2018), groups BLS
supersectors into three super-sectors:

| CORI Super-sector | Constituent BLS supersectors |
|---|---|
| **Tradable Goods** | Natural resources & mining, Construction, Manufacturing, Trade/transportation/utilities |
| **Tradable Services** | Information, Financial activities, Professional & business services |
| **Local Services** | Education & health services, Leisure & hospitality, Other services, Public administration |

Industries in the tradable categories compete in national or global markets.
Their presence in a rural county often signals competitive advantage — something
the local area does well enough that the wider market wants it. Local services,
by contrast, exist everywhere and primarily reflect local demand.

## Pulling sector data

```{r sector-data}
# CORI super-sector employment shares, county level
sector_emp <- get_sector_employment(
  geography   = "county",
  years       = 2023,
  sector_type = "CORI",
  value_type  = "share"
)

# Add rural classification
rural <- get_definition("cbsa", 2023) |> select(geoid, is_rural)

sector_emp <- sector_emp |>
  left_join(rural, by = "geoid") |>
  filter(!is.na(is_rural))

# Total employment for sizing points
total_emp <- get_employment(geography = "county", years = 2023) |>
  filter(variable == "employment") |>
  select(geoid, employment = value)
```

## Swarm plot: tradable goods share by rural status

A swarm plot shows the full distribution across counties — not just the average.
Each dot is a county, sized by employment.

```{r swarm, fig.alt = "Swarm plot showing the distribution of tradable goods employment share across rural and nonrural counties in 2023"}
swarm_data <- sector_emp |>
  filter(variable == "emp_share", sector == "tradable_goods") |>
  left_join(total_emp, by = "geoid") |>
  filter(!is.na(employment), employment > 0)

title_swarm    <- "Rural counties rely more heavily on tradable goods sectors"
subtitle_swarm <- str_wrap(
  "Share of employment in tradable goods industries (natural resources, construction, manufacturing, trade & transportation), 2023. Each dot is a county, sized by total employment.",
  width = 100
)

fig_swarm <- swarm_data |>
  ggplot(aes(
    x     = value,
    y     = is_rural,
    size  = employment,
    color = is_rural
  )) +
  geom_beeswarm(cex = 0.4, groupOnX = FALSE, alpha = 0.4) +
  scale_color_cori(palette = "ctg2ruralnonrural", guide = "none") +
  scale_size_continuous(
    range  = c(0.5, 6),
    labels = label_comma(scale = 1e-3, suffix = "k"),
    name   = "Employment"
  ) +
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_y_discrete(limits = rev(c("Rural", "Nonrural"))) +
  coord_cartesian(clip = "off") +
  theme_cori() +
  theme(
    panel.grid.major   = element_blank(),
    panel.grid.major.x = element_line(color = "#d0d2ce", linewidth = 0.5)
  ) +
  labs(
    title    = title_swarm,
    subtitle = subtitle_swarm,
    caption  = str_wrap(
      "Source: BLS QCEW via cori.data.qcew. Rural classification: CBSA 2023.",
      width = 110
    ),
    x = NULL,
    y = NULL
  )

fig_swarm
save_plot(fig_swarm, here::here("export/sector_swarm.png"))
```

## BLS supersector breakdown

For a more granular view, the 11 BLS supersectors show where employment
is concentrated across the full distribution of industry types:

```{r bls-sector}
bls_sector <- get_sector_employment(
  geography   = "county",
  years       = 2023,
  sector_type = "BLS",
  value_type  = "share"
) |>
  left_join(rural, by = "geoid") |>
  filter(!is.na(is_rural), variable == "emp_share")

# Average share by sector and rural status
bls_avg <- bls_sector |>
  left_join(total_emp, by = "geoid") |>
  group_by(sector, is_rural) |>
  summarize(
    avg_share = weighted.mean(value, employment, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r bls-bar, fig.alt = "Grouped bar chart showing employment shares by BLS supersector for rural and nonrural counties"}
fig_bls <- bls_avg |>
  ggplot(aes(
    x    = avg_share,
    y    = reorder(sector, avg_share),
    fill = is_rural
  )) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_cori(palette = "ctg2ruralnonrural") +
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  theme_cori_horizontal_bars() +
  labs(
    title    = "Sector composition differs markedly between rural and nonrural counties",
    subtitle = "Employment-weighted average sector share, 2023",
    caption  = str_wrap(
      "Source: BLS QCEW via cori.data.qcew. Rural classification: CBSA 2023.",
      width = 110
    ),
    fill = NULL,
    x    = NULL,
    y    = NULL
  )

fig_bls
save_plot(fig_bls, here::here("export/bls_sector_bars.png"))
```

## Sector wages

Pay varies substantially across sectors. Use `get_sector_wages()` to see
which sectors drive higher or lower wage outcomes:

```{r sector-wages}
sector_wages <- get_sector_wages(
  geography   = "county",
  years       = 2023,
  sector_type = "CORI"
) |>
  left_join(rural, by = "geoid") |>
  filter(!is.na(is_rural)) |>
  left_join(total_emp, by = "geoid")

# Employment-weighted average wage by sector and rural status
wage_summary <- sector_wages |>
  group_by(sector, is_rural) |>
  summarize(
    avg_wage = weighted.mean(value, agg_var, na.rm = TRUE),
    .groups = "drop"
  )
```
