---
title: "Working with Tidy Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Tidy Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{r libraries}
library(cori.data.qcew)
library(cori.data.pep)
library(cori.data.bds)
library(cori.data.bfs)
library(cori.data.bps)
library(dplyr)
library(tidyr)
library(stringr)
```

Every function in the CORI data ecosystem returns **tidy (long) format** data.
This is a deliberate design choice: tidy data pairs naturally with `dplyr`,
`ggplot2`, and the broader R data science ecosystem.

The format is always the same:

```
geoid   year  variable           value
33009   2023  employment         28450
33009   2023  job_creation        4120
33009   2022  employment         27810
...
```

One row per `geoid × year × variable`. This vignette walks through the most
common patterns for working with this format.

## Filtering by variable

The most common operation — selecting the measure you care about:

```{r filter}
bd <- get_business_dynamics(geography = "county", years = 2018:2023)

# Single variable
estabs <- bd |> filter(variable == "establishments")

# Multiple variables
rates <- bd |> filter(variable %in% c("estab_entry_rate", "estab_exit_rate"))

# Pattern match — all rate variables
all_rates <- bd |> filter(str_ends(variable, "_rate"))
```

## Pivoting to wide format

When you need variables as columns — for correlation, regression, joining
with external data, or export to a spreadsheet:

```{r pivot-wide}
bd_wide <- get_business_dynamics(geography = "county", years = 2022:2023) |>
  pivot_wider(names_from = variable, values_from = value)

# Now: geoid | year | establishments | estab_entries | estab_entry_rate | ...
glimpse(bd_wide)
```

The reverse — going from wide back to tidy — uses `pivot_longer()`:

```{r pivot-long}
# If you receive data in wide format, bring it back to tidy
bd_long <- bd_wide |>
  pivot_longer(
    cols      = -c(geoid, year),
    names_to  = "variable",
    values_to = "value"
  )
```

## Joining across packages

Because every package returns the same `geoid + year` identifiers,
joining across packages is a simple `left_join()`:

```{r join-two}
# Business applications + building permits — are they moving together?
apps <- get_business_applications(geography = "county", years = 2015:2023) |>
  filter(variable == "applications") |>
  select(geoid, year, applications = value)

permits <- get_building_permits(geography = "county", years = 2015:2023) |>
  filter(variable == "permits") |>
  select(geoid, year, permits = value)

combined <- inner_join(apps, permits, by = c("geoid", "year"))

# Correlation between applications and permits
cor(combined$applications, combined$permits, use = "complete.obs")
```

You can join any number of packages the same way. Here's a three-way join
bringing together employment, population, and business dynamics:

```{r join-three}
emp  <- get_employment(geography = "county", years = 2023) |>
  filter(variable == "employment") |>
  select(geoid, employment = value)

pop  <- get_population(geography = "county", years = 2023) |>
  filter(variable == "population") |>
  select(geoid, year, population = value)

estab <- get_business_dynamics(geography = "county", years = 2023) |>
  filter(variable == "establishments") |>
  select(geoid, establishments = value)

county_profile <- emp |>
  left_join(pop,   by = "geoid") |>
  left_join(estab, by = "geoid") |>
  mutate(
    emp_per_capita   = employment   / population,
    estab_per_capita = establishments / population
  )
```

## Filtering to specific geographies

Use `geoids` to pull specific counties or states — no filtering needed after
the fact:

```{r geoids}
# New England states (2-digit FIPS)
new_england <- c("09", "23", "25", "33", "44", "50")
ne_emp <- get_employment(geoids = new_england, years = 2018:2023)

# Appalachian counties (5-digit FIPS)
appalachian_counties <- c("54011", "54033", "21013", "21019", "47133")
app_bd <- get_business_dynamics(geoids = appalachian_counties, years = 2015:2023)
```

## Summarizing across geographies

With a rural classification joined in, aggregating to rural/nonrural
or state-level summaries is straightforward:

```{r summarize}
library(ruraldefinitions)

rural <- get_definition("cbsa", 2023) |> select(geoid, is_rural)

emp_summary <- get_employment(geography = "county", years = 2018:2023) |>
  filter(variable == "employment") |>
  left_join(rural, by = "geoid") |>
  group_by(year, is_rural) |>
  summarize(
    total_employment = sum(value, na.rm = TRUE),
    n_counties       = n_distinct(geoid),
    .groups = "drop"
  )
```

## Using ggplot2 with tidy data

Tidy data maps directly to `ggplot2` aesthetics with no reshaping:

```{r ggplot}
library(ggplot2)
library(cori.charts)
load_fonts()

# Rate trends for two variables — one call, two lines
get_business_dynamics(geography = "nation", years = 2000:2023) |>
  filter(variable %in% c("estab_entry_rate", "estab_exit_rate")) |>
  ggplot(aes(x = year, y = value, color = variable)) +
  geom_line(linewidth = 1.5) +
  scale_color_cori(
    palette = "ctg2buor",
    labels  = c("estab_entry_rate" = "Entry rate", "estab_exit_rate" = "Exit rate")
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1, scale = 1)) +
  theme_cori() +
  labs(
    title  = "National establishment entry and exit rates, 2000\u20132023",
    x      = NULL,
    y      = NULL,
    color  = NULL,
    caption = "Source: Census Bureau Business Dynamics Statistics"
  )
```
