---
title: "Weighted Averages and agg_var"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Weighted Averages and agg_var}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  eval       = FALSE,
  fig.width  = 8.888889,
  fig.height = 6.25,
  dpi        = 300
)
```

```{r libraries}
library(cori.data.qcew)
library(ruraldefinitions)
library(cori.charts)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(stringr)

load_fonts()
update_cori_geom_defaults()
```

## What is `agg_var`?

Some functions return a fifth column — `agg_var` — alongside the standard
`geoid | year | variable | value`. This is the **employment weight**: the
denominator used to properly aggregate rates and shares across geographies.

| Function | Variable | `agg_var` |
|---|---|---|
| `get_wage_salary()` | `avg_pay` | Total employment |
| `get_sector_employment()` | `emp_share` | Total employment |
| `get_sector_wages()` | `avg_pay` | Sector employment |

Functions that return raw counts (`get_employment()`,
`get_employment_concentration()`) do not include `agg_var` — counts can
simply be summed.

## The problem with simple averages

Here is the core issue. Suppose we want the average annual wage across all
rural counties in 2023.

A simple `mean()` treats every county equally — a county with 200 workers
gets exactly the same weight as one with 200,000. In practice, the result
is dominated by hundreds of small, low-employment rural counties that
collectively represent a tiny fraction of the actual rural workforce.

```{r naive-mean}
wages <- get_wage_salary(geography = "county", years = 2023) |>
  left_join(
    get_definition("cbsa", 2023) |> select(geoid, is_rural),
    by = "geoid"
  ) |>
  filter(!is.na(is_rural))

# Simple (naive) average — each county has equal weight
wages |>
  group_by(is_rural) |>
  summarize(avg_wage = mean(value, na.rm = TRUE))
```

This overstates wages in counties with large, high-paying employers (a
manufacturing plant, a hospital system) that happen to be in a small
county — and understates them in large rural counties that look more
like the typical rural experience.

## The correct approach: `weighted.mean()`

The fix is to weight each county's wage by how many workers that wage
represents. That's exactly what `agg_var` is for:

```{r weighted-mean}
wages |>
  group_by(is_rural) |>
  summarize(
    naive_avg    = mean(value, na.rm = TRUE),
    weighted_avg = weighted.mean(value, agg_var, na.rm = TRUE)
  )
```

## Visualizing the difference

```{r comparison-chart, fig.alt = "Grouped bar chart comparing simple and employment-weighted average wages for rural and nonrural counties in 2023"}
wage_comparison <- wages |>
  group_by(is_rural) |>
  summarize(
    `Simple average`        = mean(value, na.rm = TRUE),
    `Employment-weighted`   = weighted.mean(value, agg_var, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols      = -is_rural,
    names_to  = "method",
    values_to = "avg_wage"
  )

title    <- "Employment-weighting changes the picture on rural wages"
subtitle <- "Average annual pay per worker, 2023"
caption  <- str_wrap(
  "Source: BLS Quarterly Census of Employment and Wages via cori.data.qcew. Rural classification: CBSA 2023.",
  width = 110
)

fig_bar <- wage_comparison |>
  ggplot(aes(x = is_rural, y = avg_wage, fill = method)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(
    aes(label = label_dollar(accuracy = 1)(avg_wage)),
    position  = position_dodge(width = 0.7),
    vjust     = -0.5,
    size      = 3.5,
    family    = "Lato",
    fontface  = "bold"
  ) +
  scale_fill_cori(palette = "ctg2buor") +
  scale_y_continuous(
    labels = label_dollar(scale = 1e-3, suffix = "k"),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.15))
  ) +
  theme_cori() +
  labs(
    title    = title,
    subtitle = subtitle,
    caption  = caption,
    fill     = NULL,
    x        = NULL,
    y        = NULL
  )

fig_bar
save_plot(fig_bar, here::here("export/wage_weighted_comparison.png"))
```

## Aggregating across geographies

`agg_var` also enables correct aggregation when you want to roll up counties
into regions. Summing wages is meaningless — you need to weight by employment:

```{r aggregate-regions}
library(ruraldefinitions)

rural <- get_definition("cbsa", 2023) |> select(geoid, is_rural)

# Correct: employment-weighted regional average
sector_wages <- get_sector_wages(
  geography   = "county",
  years       = 2023,
  sector_type = "CORI"
) |>
  left_join(rural, by = "geoid") |>
  filter(!is.na(is_rural))

regional_wages <- sector_wages |>
  group_by(sector, is_rural) |>
  summarize(
    # agg_var here is sector employment — the correct weight for sector wages
    regional_avg_wage = weighted.mean(value, agg_var, na.rm = TRUE),
    total_sector_emp  = sum(agg_var, na.rm = TRUE),
    .groups = "drop"
  )
```

## Rule of thumb

Whenever a function returns `agg_var`, use `weighted.mean(value, agg_var)`
instead of `mean(value)` any time you're aggregating across multiple geographies.
The difference matters most in rural analysis, where high variance in county
size means naive averages can mislead significantly.
